---
title: "Noncollapsibility of the odds ratio"
date: "2023-12-02"
description: "IPW-analysis"
categories: [code, analysis]

bibliography: 
  - "/home/meichtry/Templates/Rcite.bib"
  - grateful-refs.bib
---






# Noncollapsibility

We simulate data from GLM with **non-confounding** covariate $C$:



```{r }
set.seed(3)
N <- 400
C <- sort(runif(N,-8,8)) #non-confounding C
G<-sample(c(0,1),N,replace=TRUE)
beta0<-0
beta1<-log(10)
beta2<-1
##Logistic 
etai<-beta0+beta1*G+beta2*C
pii<-exp(etai)/(1+exp(etai))
Ydich <-rbinom(G,size=1,prob=pii)
dat<-data.frame(C,G,Ydich)
cor(dat)
```


```{r }
modC<-glm(Ydich~G+C,family="binomial")
modM<-glm(Ydich~G,family="binomial")
```



## Illustrate non-collapsibility

```{r }
pred<-predict(modM,type="response",se.fit=TRUE)
predgA<-predict(modC,newdata=data.frame(C=C[G==0],G=0),se.fit=TRUE,type="response")
predgB<-predict(modC,newdata=data.frame(C=C[G==1],G=1),se.fit=TRUE,type="response")
plot(C,Ydich,col=c("blue","red")[G+1],ylab="Probability")
lines(sort(C[G==0]),sort(predgA$fit),col="blue")
lines(sort(C[G==1]),sort(predgB$fit),col="red")
abline(a=mean(pii[G==0]),b=0,col="blue",lty=2)
abline(a=mean(pii[G==1]),b=0,col="red",lty=2)
```

## Marginal model



```{r }
summary(modM)$coef
```

## Conditional model

Effect of $G$ is larger, even though $C$ is not a confounder. This is due to **non-collapsibility of the odds ratio**


```{r }
summary(modC)$coef
```


## IPW-model

Estimate inverse probability weights to fit marginal structural
     models in a point treatment situation. 


### Weights


```{r }
library(ipw)
temp <- ipwpoint(exposure=G,family = "binomial",link = "logit",data=dat,numerator =~1,denominator = ~ C)
summary(temp$ipw.weights)
dat$sw<-temp$ipw.weights
```

### Marginal structural model

Marginal structural model for the causal effect of $X$ on $Y$ corrected for confounding by $C$ using inverse probability weighting

```{r }
modW<-glm(Ydich~G,weights=sw,data=dat,family=quasibinomial)
summary(modW)$coef
```


```{r eval=FALSE}
require("survey")
msm <- (svyglm(Ydich ~ G, family="binomial",design = svydesign(~ 1, weights =~sw,data = dat)))
summary(msm)
```


## Comparison


```{r }
modelsummary::modelsummary(models=list("marginal"=modM,"conditional"=modC,"IPW"=modW))
```

::: {.callout-important}
* conditional-marginal=noncollaps+confouding
* IPW-marginal=confounding
* conditional-IPW=noncollaps
:::




# Example smoking and children weight

```{r echo=FALSE}
library(rio)
bwt <- import("bwt_full.csv")
#bwt$low <- factor(bwt$low, levels = c(0,1), labels = c("no", "yes"))
#bwt$smoke <- factor(bwt$smoke, levels = c(0,1), labels = c("no", "yes"))
bwt$socclass <- factor(bwt$socclass, levels = c("I", "II", "III"), labels = c("I", "II", "III"))
```
```{r }
str(bwt)
```

```{r }
glm_smoke <- glm(low ~ smoke, data = bwt, family = binomial)
summary(glm_smoke)$coef
glm_multi_1 <- glm(low ~ smoke + socclass, data = bwt, family = binomial)
summary(glm_multi_1)$coef
```


```{r }
temp <- ipwpoint(exposure=smoke,family = "binomial",link = "logit",data=bwt,
                 numerator=~1,denominator=~socclass)
summary(temp$ipw.weights)
bwt$sw<-temp$ipw.weights
```

```{r }
modW<-glm(low~smoke,weights=sw,data=bwt,family=quasibinomial)
summary(modW)$coef
```


::: {.callout-important}
* conditional-marginal=noncollaps+confouding
* IPW-marginal=confounding
* conditional-IPW=noncollaps
:::


```{r }
modelsummary::modelsummary(models=list("marginal"=glm_smoke,"conditional"=glm_multi_1,"IPW"=modW))
```


# Potential outcomes
See [Example](https://www.r-bloggers.com/2017/12/when-you-use-inverse-probability-weighting-for-estimation-what-are-the-weights-actually-doing/) for marginal structural models in this context.

## Simulate some data

```{r }
library(simstudy)
# define the data
defB <- defData(varname = "L", formula =0.27, 
                dist = "binary")
defB <- defData(defB, varname = "Y0", formula = "-2.5 + 1.75*L", 
                dist = "binary", link = "logit")
defB <- defData(defB, varname = "Y1", formula = "-1.5 + 1.75*L", 
                dist = "binary", link = "logit")
defB <- defData(defB, varname = "A", formula = "0.315 + 0.352 * L", 
                dist = "binary")
defB <- defData(defB, varname = "Y", formula = "Y0 + A * (Y1 - Y0)", 
                dist = "nonrandom")
defB
```

```{r }
set.seed(2002)
Nnew<-1000
dtB <- genData(Nnew, defB)
dtB
```

## True causal effect (based on potential outcomes)


```{r }
odds <- function (p) {
    return((p/(1 - p)))
}

dtB[, log( odds( mean(Y1) ) / odds( mean(Y0) ) )]
```

## Conditional effect

The true conditional causal effect of $A$ is 1.

```{r }
mc<-glm(Y ~ A + L , data = dtB, family="binomial")
mc
```
This estimate for $A$ is a good estimate of the conditional effect in the population, based on the potential outcomes at each level of $L$

```{r }
dtB[, .(LOR = log( odds( mean(Y1) ) / odds( mean(Y0) ) ) ), keyby = L]
```

## Marginal effect

The marginal estimate is biased both for the conditional effect and the marginal causal effect.

```{r }
mm<-glm(Y ~ A , data = dtB, family="binomial")
mm
```


## Numerator and Denominator model

```{r }
numModel <- glm(A ~ 1, data = dtB, family = "binomial")
denModel <- glm(A ~ L, data = dtB, family = "binomial")
```

## Stabilized weights by hand

```{r }
dtB[, pA0 := predict(numModel, type = "response")]
dtB[, pA := predict(denModel, type = "response")]
defB2 <- defDataAdd(varname = "IPW", 
                    formula = "(A*pA0+(1-A)*(1-pA0))/((A * pA) + ((1 - A) * (1 - pA)))", 
                    dist = "nonrandom")
dtB <- addColumns(defB2, dtB)
dtB[1:6]
```

```{r }
unique(dtB$IPW)
```

## Stabilized weights with **ipw** package

```{r }
tempsw<-ipw::ipwpoint(exposure=A,family="binomial",link="logit",numerator=~1,denominator = ~L,data=dtB)
tempw<-ipw::ipwpoint(exposure=A,family="binomial",link="logit",denominator = ~L,data=dtB)
unique(tempsw$ipw.weights)
mean(tempsw$ipw.weights)
unique(tempw$ipw.weights)
mean(tempw$ipw.weights)
```


## Applying IPW 

### Unstabilized weights

```{r }
summary(mw<-glm(Y ~ A , data = dtB, family="binomial", weights = tempw$ipw.weights))$coef
```

### Stabilized weights

```{r }
summary(mws<-glm(Y ~ A , data = dtB, family="binomial", weights = tempsw$ipw.weights))$coef
```
## Comparison

```{r }
modelsummary::modelsummary(models=list("marginal"=mm,"conditional"=mc,"IPW"=mw,"IPWs"=mws))
```

```{r echo=FALSE}
library(grateful)
cite_packages(pkgs = "Session",output="paragraph",out.dir=getwd(),cite.tidyverse=FALSE,omit=c("grateful","knitr"))
cite_packages(pkgs = "Session",output="table",out.dir=getwd(),cite.tidyverse=FALSE,omit=c("grateful","knitr"))
```


```{r EffSampleSize, include=FALSE}
Nnew
w<-tempw$ipw.weights
sw<-tempsw$ipw.weights
sum(w)
sum(sw)
sum(w)^2/sum(w^2)
sum(sw)^2/sum(sw^2)
Nnew/(1+var(w))
Nnew/(1+var(sw))
```



# Bibliography {-}












