---
title: "Compositions"
date: "2022-11-03"
description: "Simplex as sample space: Tetrahedron for 4-part compositions"
categories: [code, analysis]
---


###### Sample Space of compositional data ######

The sample space of compositional data with $D$ parts is a simplex $\mathcal{S}^D$ of dimension $D-1$: 

$\mathcal{S}^D=\{\mathbf{x}=[x_1,x_2,\dots,x_D]\in \mathbb{R}^D \mid
x_i>0, i=1,2,\dots,D; \sum_{i=1}^Dx_i=\mathcal{k}\}$

The only information is given by the *ratios between components*, so
the information of a composition is preserved under multiplication by
any positive constant. Therefore, the sample space of compositional
data can always be assumed to be a standard simplex, i.e. $\mathcal{k}
= 1$.

Normalization to the standard simplex is called closure and is

$\mathcal{C}[x_1,x_2,\dots,x_D]=\left[\frac{x_1}{\sum_{i=1}^D x_i},\frac{x_2}{\sum_{i=1}^D x_i}, \dots,\frac{x_D}{\sum_{i=1}^D x_i}\right]$


###### Example ######

4 Raters rated a 4-part-composition on $n=20$ subjects. See the interactive *Tetrahedron*

```{r include=FALSE,warning=FALSE}
library(compositions)
library(DirichletReg)
library(psych)
library(irr)
library(rgl)
knitr::knit_hooks$set(webgl = hook_webgl)
d.ern<-read.csv("ern.csv",stringsAsFactors=TRUE)
d.ern[]<-lapply(d.ern,gsub,pattern="%",replacement="")
d.ern[]<-lapply(d.ern,as.numeric)
str(d.ern)
d.ern
d.ern$ID<-as.factor(d.ern$ID)
data.frame(names(d.ern))
A<-acomp(d.ern[,c(2,6,10,14)])
B<-acomp(d.ern[,c(3,7,11,15)])
C<-acomp(d.ern[,c(4,8,12,16)])
D<-acomp(d.ern[,c(5,9,13,17)])

HA<-as.matrix(d.ern[,c(2,6,10,14)])%*%c(0,1,2,3)
HB<-as.matrix(d.ern[,c(3,7,11,15)])%*%c(0,1,2,3)
HC<-as.matrix(d.ern[,c(4,8,12,16)])%*%c(0,1,2,3)
HD<-as.matrix(d.ern[,c(5,9,13,17)])%*%c(0,1,2,3)

apply(A,1,sum) 
apply(B,1,sum)
apply(C,1,sum) 
apply(D,1,sum)
datAitch<-acomp(data.frame(rbind(A,B,C,D)))
datAitch
Rater<-as.factor(rep(c("A","B","C","D"),each=20))
ID<-as.factor(rep(1:20,4))
peri<-d.ern[,2:5]
zent<-d.ern[,6:9]
kort<-d.ern[,10:13]
veg<-d.ern[,14:17]
```



```{r  test-rgl,webgl=TRUE,warning=FALSE}
plot3D(datAitch,col=as.numeric(Rater),size=5,axes=TRUE)
```
